\chapter{Conclusion} \label{sec:conclusion}

Having thoroughly scrutinized the theoretical basis of evolvability, we will conclude by using evolvability as a lens to reflect on the relation between evolutionary biology and evolutionary computing to develop a broader perspective on the topic. 
It may not be surprising that evolvability is a topic of active discussion among elements of the evolutionary biology community \cite{Pigliucci2008IsEvolvable}.
The concept falls under the umbrella of a broader effort to expand the theoretical framework of evolution called the extended evolutionary synthesis \cite{Pigliucci2007DoSynthesis}.
Compared to the EA community, however, the concept of evolvability has been slower to gain traction among evolutionary biologists.
As Kirschner and Gerhart, a pair of biologists known for their theory of facilitated variation, comment,
\begin{displayquote}
``Many evolutionary biologists do not see a need to connect somatic adaptability to the generation of variation, and some see a need to keep them separate. For them, it is sufficient to say that random mutation is required and that the phenotypic variation arises haphazardly from it as random damage; the organism's current phenotype does not matter for the variation produced, and the output of variation is nearly random \cite[p 219]{Kirschner2005TheDilemma}.''
\end{displayquote}
Perhaps, in part, evolutionary biologists are less predisposed to interest in evolvability because they are not so directly stymied its absence.
Success in attempts to emulate the evolutionary process to generate designs for sophisticated systems such as artificial neural networks or robotic bodies hinges on the ability of the evolutionary algorithm to generate viable, heritable variation.
This development of this capability has been a major hurdle in EA research, especially in the field's early years.
The intense and ubiquitous interest in evolvability among the EA community should therefore come as no surprise.

EAs have yielded interesting and useful results, but have not yet come close to replicating the intricacy or scale of biological systems \cite{Tonelli2011OnSystems}.
In classical EAs, a directly or trivially indirectly encoded population evolves against a static fitness function.
Fitness gains are typically realized for a period of several hundred generations before innovation stagnates and the population settles out at an equilibrium.
This approach, predicated on a fundamentally accurate but extremely simplistic view of evolution, yields limited results.
The stunted effectiveness of early EAs can be cast as a reflection of the limitations innate to the theory on which the algorithms are built.
EA research provides uniquely direct and blunt evidence that it is \textit{not} sufficient to say that ``phenotypic variation arises haphazardly'' \cite[p 219]{Kirschner2005TheDilemma}.
EA can lend ammunition to biologists who posit that the ability of organisms to yield offspring more fit than themselves through ``mutation, recombination and development'' is ``surprising and... demands an explanation'' \cite{Draghi2008EVOLUTIONMODEL}.
As skeptics in the evolutionary biology community point out, ``the best way to elevate the prominence of genuinely interesting phenomena... is to strengthen the evidence for their importance'' \cite{Laland2014DoesRethink}.
The evolutionary computing community have enjoyed a fruitful thrust to develop research algorithms that incorporate a broader array of theoretical factors that may influence evolution, such as varying fitness functions and phenotypic plasticity \cite{Kashtan2007VaryingEvolution,Moczek2011TheInnovation,Downing2012HeterochronousBaldwinism}. Perhaps the evolutionary biology community would find widespread elevation of such theory beyond the status of circumstantial ``‘add-ons’ to the basic processes that produce evolutionary change'' similarly fruitful \cite{Laland2014DoesRethink}. 

At present, it seems likely that evolvability stems from a large and diffuse web of cooperating mechanisms.
The establishment --- or rejection --- of empirical evidence for causal links between factors such as plasticity or the developmental process and evolvability must be a key research goal in the field of evolutionary algorithm design.
Such results will directly support efforts to refine the evolutionary algorithm and realize performance more closely akin to that of its biological counterpart.
This line of inquiry raises and addresses questions of interest to the evolutionary biology community, especially in light of continuing controversy surrounding the extensions to the evolutionary synthesis.
It will help determine which theoretical elaborations are necessary to account for evolution as observed in biology.
It is hoped that further research in this vein --- both \textit{in silico} and \textit{in vivo} --- and, especially, continued exchange between EA and evolutionary biology researchers will yield both biological insight and more powerful digital engineering techniques.